The topos-theoretic multiverse: \\ a modal approach for computation

### 2024-09-27 paper-modal-multiverse.tex
to do
Sync with Overleaf version

\newcommand{\xra}[1]{\xrightarrow{#1}}
\newcommand{\XXX}[1]{\textbf{\textcolor{red}{XXX: #1}}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\nnn}{\mathfrak{n}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\U}{\mathcal{U}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\QQ}{\mathbb{Q}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Mod}{Mod}
\DeclareMathOperator{\Sh}{Sh}
\DeclareMathOperator{\PSh}{PSh}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\renewcommand{\_}{\mathpunct{.}\,}
\newcommand{\effective}{ef{}fective\xspace}
\newcommand{\notnot}{\emph{not~not}\xspace}

@@

@@

@@

\noindent
Thanks to the finer distinctions constructive mathematics offers, there is a
host of principles which are available in classical mathematics but seem naive
from a constructive point of view. A non-exhaustive list is:
\begin{enumerate}
\renewcommand{\theenumi}{\arabic{enumi}*}
\item A transitive relation is well-founded iff there is no infinite descending
chain.
\item A relation is almost-full iff every infinite sequence is good (meaning
that some term on the sequence is related to some later term -- the notion
of almost-full relations
has been studied in combinatorics~\cite{xxx} and found applications in
termination checking~\cite{xxx}).
\item \emph{Krull's lemma}: A ring element is nilpotent iff every prime
ideal contains it.
\item Every ring has a maximal ideal.\footnote{Here and in the following, by
\emph{ring} we mean commutative ring with unit and by \emph{maximal ideal} we
mean an ideal~$\mmm$ which is \emph{proper} in the sense that~$1 \in \mmm
\Rightarrow 1 = 0$ and such that for every proper ideal~$\nnn$ with~$\mmm
\subseteq \nnn$, $\mmm = \nnn$.}
\item \emph{Markov's principle:} If a function~$\NN \to \NN$ does \notnot have
a zero, then it actually has a zero.
\item \emph{Dependent choice:} If every element of a set is related by some relation to some other
element, then every element can be completed to an infinite chain of related
elements.
\item The law of excluded middle holds.
\end{enumerate}
Constructive theorems always carry computational and geometric
content---from every constructive proof, a corresponding algorithm can be
extracted~\cite{bauer:c2c}, and every constructive proof holds also for continuous families of
the objects in question~\cite[Section~4.3]{blechschmidt:filmat}. In contrast, listed classical principles above have no
computational witness and/or fail in continuous families, hence are not
available in constructive mathematics.

@@

Briefly, a statement~$\varphi$ is said to hold \emph{everywhere}
($\necessary\varphi$) iff it holds in every Grothendieck topos over the
current base topos (makng use of the \emph{internal language} of
toposes~\cite{xxx}, where required in the form of the \emph{sheaf semantics} in
order to support unbounded quantification~\cite{xxx}); a statement holds \emph{somewhere} ($\possible\varphi$)
iff it holds in some positive Grothendieck topos over the current base; and a
statement holds \emph{proximally} ($\xpossible\varphi$) iff it holds in
some positive ouvert Grothendieck topos over the current base. More such
\emph{modalities} are also useful and merit study; the precise definitions are
given in Section~\ref{sect:defn-modalities}.

This modal language not only allows us to recover classical principles
as above, but also makes some powerful theorems about the topos-theoretic
landscape smoothly accessible:
\begin{enumerate}
\addtocounter{enumi}{7}
\item \emph{Barr's theorem, full version:} If Zorn's lemma holds, it is
everywhere the case that it (and even the full axiom of choice) hold somewhere.
\item If a geometric implication holds \emph{somewhere}, then it holds already here.
\item \label{item:prox-descends}
If a first-order statement holds \emph{proximally}, then it holds already here.
\item \label{item:prox-countable}
For every (perhaps uncountable) inhabited set~$X$, \emph{proximally} there is a
surjection~$\NN \twoheadrightarrow X$.
\end{enumerate}
An example application of the latter two principles
has recently been studied in constructive
commutative algebra~\cite{blechschmidt-schuster:constructive-maximal-ideals}.
For countable rings, an explicit construction of a maximal ideal is available.
By Item~(\ref{item:prox-countable}), this construction can also be carried out
for arbitrary rings, though the result is not a maximal ideal in the narrow
sense; rather, the resulting maximal ideal exists \emph{proximally}, in some
positive ouvert Grothendieck topos. However, the first-order consequences of
its existence, pertaining for instance to concrete statements about polynomials
or matrices, pass down to the base topos by Item~(\ref{item:prox-descends}).
The resulting maximal ideal can be regarded as a mathematical
phantom in the sense of Gavin Wraith~\cite{xxx}, not existing in a narrow
sense (in the base topos), but existing proximally and hence encouraging us to broaden our notion of
existence because it promises us to work wonders.

We can also adopt the notions of \emph{switches} and \emph{buttons} from the
modal study of the set-theoretic multiverse~\cite[xxx]{xxx}. Switches are
statements~$\varphi$ such that~$\necessary(\possible\varphi \wedge
\possible\neg\varphi)$ and buttons are statements~$\varphi$ such
that~$\necessary\possible\necessary\varphi$; switches can be toggled on and off
like a light switch, while buttons once pressed cannot be unpressed:
\begin{enumerate}
\addtocounter{enumi}{10}
\item \emph{The law of excluded middle is a switch:} Everywhere it is the case
that somewhere~\textsc{lem} holds and somewhere it does not.
\item \emph{Being countable is a button:} For every set~$X$, everywhere it is
the case that somewhere (even proximally) it is case that everywhere~$X$ is
countable.
\end{enumerate}

We argue that the modal operators~$\necessary,\possible,\xpossible$ and more
suggested are natural extensions and refinements of the
familiar double-negation modality~$\neg\neg$ in constructive mathematics.

\textbf{Acknowledgments.} Alexander Oldenziel, \ldots
% xxx: Iosif

xxx: idea of multiverse not new, but arguably basic to topos theory; new is systematic
study of its modal nature with the focus on applications

xxx: related work: Hamkins, Victoria Gitman, ... concentrates on exploring
range of set-theoretic possibility, obtaining beautiful results such as ...
Shawn Henry, Andreas Blass ...

xxx: This text is set in the context of constructive mathematics.
Our definitions and results can be formalized in~\textsc{izf} or in the
kind of language supported by toposes.

@@

Henceworth, by the unqualified word ``topos'' we will mean ``Grothendieck topos
over the current base'', and the current base topos will be denoted~``$\Set$''.
The base topos might be the ``true category of sets'', assuming that this
concept is available in one's ontology, or also some other elementary topos
such as the free topos~\cite{xxx} or the ef{}fective topos~\cite{xxx}.

The xxx (use word of Hamkins) of a base~$\B$ is the collection of Grothendieck
toposes over~$\B$ (which in set-theoretic foundations should more precisely be
formalized as the proper class of small sites in~$\B$).

\begin{remark}Perhaps arbitrary \emph{elementary} toposes over the base, corresponding to
possibly unbounded geometric morphisms~$\E \to \B$, or even arbitrary
fibrations/indexed categories over~$\B$ (XXX: are these the same?) validating an appropriate form of the
axioms of elementary toposes, should also be
taken as part of the topos-theoretic multiverse. The first foray into the
modal topos-theoretic multiverse outlined in this note
sticks to Grothendieck toposes for ease of
formalizability (``for every small site'' can be expressed also in the more
standard flavors of the internal topos language, while ``for every elementary
topos'' requires more elaborate versions); because the restricted multiverse
is already sufficiently rich for the intended applications; and because we have
a generalization to the predicative setting~\cite{crosilla:xxx} with arithmetic
universes in mind. Predicatively, not even the category of sets might be an
elementary topos, but the category of sets and categories of sheaves are still
arithmetic universes.
\end{remark}

@@

\begin{definition}A topos~$\E$ is \emph{positive} if and only if the geometric
morphism~$f : \E \to \Set$ is surjective (that is~$f^*$ reflects
isomorphisms).\end{definition}

@@

\begin{example}A sufficient criterion for the classifying topos of a geometric
theory~$\TT$ to be positive is that xxx.\end{example}

\subsection{Ouvert toposes}

\begin{definition}A topos~$\E$ is \emph{ouvert} if and only if the geometric
morphism~$f : \E \to \Set$ is open, that is~$f^*$ preserves the interpretation of
first-order formulas.\end{definition}

\begin{example}The topos of sheaves over a topological space is always ouvert. More
generally, the topos of sheaves over a locale~$X$ is ouvert iff there exists a
\emph{positivity predicate} on its frame of opens in the sense of~\cite{xxx}.
With~\textsc{lem}, every locale is ouvert~\cite{xxx}.\end{example}

\begin{example}The spectrum of a ring~$A$ is ouvert iff every element of~$A$ is
nilpotent or not~\cite[Proposition~12.51]{blechschmidt:phd}.\end{example}

\begin{example}A sufficient criterion for the classifying topos of a geometric
theory~$\TT$ being ouvert is that the indexing sets of all disjunctions
appearing on the right hand side of turnstiles, in a normal form presentation
of~$\TT$, are inhabited~\cite{xxx}.\end{example}

@@

xxx: properly define language

@@

Similar inductive notions are used to reformulate other classical definitions
in a constructively more sensible way. For instance, the classical definition
of a binary relation~$R$ on a set~$X$ being \emph{almost-full} is ``every
infinite sequence~$\alpha : \NN \to X$ is \emph{good} in the sense that there
exist numbers~$i < j$ such that~$\alpha(i) \mathrel{R} \alpha(j)$''.
For a constructive reformulation, we shift to finite approximations of
infinite sequences (finite lists of elements of~$X$) and define when such an
approximation is deemed good:\footnote{By~``$\sigma[n]$'', we mean the element
at position~$n$ of the finite list~$\sigma$. This notation is only meaningful
if the length of~$\sigma$ is at least~$n+1$. By~``$\sigma :: x$'' we mean the
enlarged list which has~$x$ as an extra element at its tail end, and by~``$[]$'' we
denote the empty list. In computer science practice, it is often more efficient
to prepend (``$x :: \sigma$'') instead of append, but this detail shall not
concern us here.}
\[ \mathsf{Good}(\sigma) \defeqv (\exists i<j\_ \sigma[i] \mathrel{R} \sigma[j]). \]
We then inductively generate a relation~``$P \,|\, \sigma$'' for monotonous
predicates~$P$ on finite lists expressing that no matter how the given finite
approximation~$\sigma$ evolves over time to a better approximation,
eventually~$P$ will hold, by the clauses
\[
  \infer{P \,|\, \sigma}{P(\sigma)}
  \qquad
  \infer{P \,|\, \sigma}{\forall x \? X\_\ P \,|\, (\sigma :: x)}
\]
and finally define that~$R$ is almost-full iff~``$\mathsf{Good} \,|\, []$''.
With this inductive definition, expected properties of the class of almost-full
relations such as stability under cartesian products (Dickman's lemma),
finite lists (Higman's lemma) or finitely-branching trees (Kruskal's theorem)
can all be constructively verified~\cite{xxx}.

A similar such definition has been proposed by XXX and XXX in commutative
algebra for expressing that a ring is Noetherian~\cite{xxx}; the classical
definition ``every ascending chain of ideals stabilizes'' and also the more
meaningful and classically equivalent characterization as ``every ascending
chain of finitely generated ideals stalls''\footnote{A chain~$\aaa_0 \subseteq
\aaa_1 \subseteq \cdots$ \emph{stalls} iff for some number~$n \in \NN$,
$\aaa_{n+1} = \aaa_n$. We are grateful to Matthias Hutzler for proposing this
terminology.} are constructively too weak; firstly, without the axiom of
dependent choice we can often not construct such chains~\cite{richman:xxx} (but only
``multi-valued chains'' as in~\cite[Section~3.9]{blechschmidt:phd}; but also
see~\cite{richman:xxx}), and secondly, being able to inspect suitable inductive
witnesses enables us to prove the Hilbert basis theorem~\cite{xxx}. XXX and XXX
hence propose to call a ring \emph{Noetherian} if and only if~$\mathsf{Stalls}
\,|\, []$, where~$\mathsf{Stalls}$ is the predicate on finite lists of finitely
generated ideals expressing xxx.

@@

\begin{proof}The classifying topos of the theory of an infinite sequence
over~$X$ can be presented as the topos of sheaves over the site given by the
partially ordered set of finite lists of elements of~$X$ with coverage given
by xxx (see Appendix xxx). The Kripke--Joyal semantics states that the
statement ``$\alpha_0$ is good'', where~$\alpha_0$ is the generic infinite
sequence, holds in the classifying topos if and only if there is a
covering~$\U$ of~$[]$ such that for every open~$U \in \U$, xxx. This precisely
amounts to~$R$ being almost-full in the inductive sense.
\end{proof}

@@

\begin{proof}Similar as the proof of Proposition~\ref{prop:gen-good}. Details
for the variant of ``bad sets'' instead of ``infinite descending chains'' have
been developed (xxx:language) by xxx~\cite{xxx}.\end{proof}

@@

    \item By \textsc{lem} and well-foundedness, there is a minimal value~$\alpha(i_0)$
    among all values of~$\alpha$.
    Similarly, there is a minimal value~$\alpha(i_1)$ among~$(\alpha(n))_{n >
    i_0}$, a minimal value~$\alpha(i_2)$ among~$(\alpha(n))_{n > i_1}$, and so
    on. By \emph{proximal dependent choice}, we can proximally collect these indices
    into a function~$i : \NN \to \NN$;\footnote{We could also avoid dependent
    choice by picking not some suitable indices~$i_0, i_1, \ldots$, but the
    least such.} this switches~\textsc{lem} off.

    \item Switching~\textsc{lem} on again, there is a minimal value~$\beta(i(k_0))$
    among all values of~$\beta \circ i$. Hence~$\gamma$ is good in view of
    \begin{align*}
      \alpha(i(k_0)) &\leq \alpha(i(k_0+1)), &
      \beta(i(k_0)) &\leq \beta(i(k_0+1)).
    \end{align*}
  \end{enumerate}
\end{proof}

@@

xxx: determine modal laws

xxx: more case studies

xxx: don't forget right adjoints

xxx: predicative

\bibliographystyle{splncs04}
\bibliography{mapaper-modal-multiverse}

@@

>>>>

\newcommand{\xra}[1]{\xrightarrow{#1}}
\newcommand{\xxx}[1]{}
\newcommand{\XXX}[1]{\textbf{\textcolor{red}{XXX: #1}}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\nnn}{\mathfrak{n}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\U}{\mathcal{U}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\QQ}{\mathbb{Q}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Mod}{Mod}
\DeclareMathOperator{\Sh}{Sh}
\DeclareMathOperator{\PSh}{PSh}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\renewcommand{\_}{\mathpunct{.}\,}
\newcommand{\effective}{ef{}fective\xspace}
\newcommand{\notnot}{\emph{not~not}\xspace}

@@

\author{Ingo Blechschmidt\inst{1} \and Alexander Gietelink Oldenziel\inst{2}}
\institute{Arberstr. 5, 86159 Augsburg, Germany
\email{iblech@speicherleck.de}
\and University College London, Gower Street, London, WC1E 6BT
\email{a.f.d.a.gietelinkoldenziel@gmail.com}}

@@

\xxx{explain somewhere that just as $\neg\neg$ is visible constructively but not classically and that $\neg\neg$ enriches our understanding, so do the new modal operators. Especially in view of the fact that if LEM holds, then $\varphi \Leftrightarrow \necessary\varphi$ for bounded first-order formulas, as in the presence of LEM every topos is overt.}

\xxx{state (and ponder more) the example that every field has a separable closure somewhere}

\xxx{Alexander:
- it is not the case that for every topos E and every field k in E there is a separable closure k_s / k i don't think. what is a counterexample?
- it should be the case that for every (geometric field?)
-> it is positive (since it the etale topos of a field is Galois over the base so the pullback reflects isomorphisms I think: the pullback is a constant sheaf functor which sends a set to the trivial G-set on that set.)}

@@

\noindent
Thanks to the finer distinctions constructive mathematics offers, there is a
host of principles which are available in classical mathematics but seem naive
from a constructive point of view. A non-exhaustive list is:
\begin{enumerate}
\renewcommand{\theenumi}{\arabic{enumi}*}
\item A transitive relation is well-founded iff there is no infinite descending
chain.
\item A relation is almost-full iff every infinite sequence is good.\footnote{An infinite sequence is \emph{good} iff some term of the sequence is related to some later term. The notion
of almost-full relations
has been studied in combinatorics~\cite{vytiniotis-coquand-wahlstedt:af,carroy-pequignot:well,lescanne:well} and found applications in
termination checking~\cite{blass-gurevich:well}.}
\item \emph{Krull's lemma}: A ring element is nilpotent iff every prime
ideal contains it.
\item Every ring has a maximal ideal.\footnote{Here and in the following, by
\emph{ring} we mean commutative ring with unit and by \emph{maximal ideal} we
mean an ideal~$\mmm$ which is \emph{proper} in the sense that~$1 \in \mmm
\Rightarrow 1 = 0$ and such that for every proper ideal~$\nnn$ with~$\mmm
\subseteq \nnn$, $\mmm = \nnn$.}
\item \emph{Markov's principle:} If a function~$\NN \to \NN$ does \notnot have
a zero, then it actually has a zero.
\item \emph{Dependent choice:} If every element of a set is related by some relation to some other
element, then every element can be completed to an infinite chain of related
elements.
\item The law of excluded middle holds.
\end{enumerate}
Constructive theorems always carry computational and geometric
content---from every constructive proof, a corresponding algorithm can be
extracted~\cite{bauer:c2c}, and every constructive proof holds also for continuous families of
the objects in question~\cite[Section~4.3]{blechschmidt:filmat}. In contrast, the listed classical principles above have no
computational witness and/or fail in continuous families, hence are not
available in constructive mathematics.

@@

Briefly, a statement~$\varphi$ is said to hold \emph{everywhere}
($\necessary\varphi$) iff it holds in every Grothendieck topos over the
current base topos (making use of the \emph{internal language} of
toposes~\cite{borceux:handbook3,streicher:ctcl,maietti:modular});\footnote{In order to support unbounded quantification we occasionally make use of the extension of the usual Kripke--Joyal semantics in the form of \emph{stack semantics}~\cite{shulman:stack-semantics})} a statement holds \emph{somewhere} ($\possible\varphi$)
iff it holds in some positive Grothendieck topos over the current base; and a
statement holds \emph{proximally} ($\xpossible\varphi$) iff it holds in
some positive overt Grothendieck topos over the current base. More such
\emph{modalities} are also useful and merit study; the precise definitions are
given in Section~\ref{sect:defn-modalities}.

This modal language not only allows us to recover classical principles
as above, but also makes some powerful theorems about the topos-theoretic
landscape smoothly accessible:
\begin{enumerate}
\addtocounter{enumi}{7}
\item \emph{Barr's theorem, full version:} If Zorn's lemma holds, it is
everywhere the case that it (and even the full axiom of choice) hold somewhere.
\item If a geometric sequent $\wedge_{i=1}^n \phi_i \vdash \bigvee_{j\in J}\psi_j$ holds \emph{somewhere}, then it holds already here.
\item \label{item:prox-descends}
If a bounded first-order statement holds \emph{proximally}, then it holds already here.
\item \label{item:prox-countable}
For every (perhaps uncountable) inhabited set~$X$, \emph{proximally} there is a
surjection~$\NN \twoheadrightarrow X$.
\end{enumerate}
An example application of the latter two principles
has recently been studied in constructive
commutative algebra~\cite{blechschmidt-schuster:constructive-maximal-ideals}:
For countable rings, an explicit iterative construction of a maximal ideal is available.
By Item~\ref{item:prox-countable}, this construction can also be carried out
for arbitrary rings, though the result is not a maximal ideal in the narrow
sense; rather, the resulting maximal ideal exists \emph{proximally}, in some
positive overt Grothendieck topos. However, the first-order consequences of
its existence, pertaining for instance to concrete statements about polynomials
or matrices, pass down to the base topos by Item~\ref{item:prox-descends}.
The resulting maximal ideal can be regarded as a mathematical
phantom in the sense of Gavin Wraith~\cite{wraith:phantoms}, not existing in a narrow
sense (in the base topos), but existing proximally and hence encouraging us to broaden our notion of
existence because it promises us to work wonders.

We can also adopt the notions of \emph{switches} and \emph{buttons} from the
modal study of the set-theoretic multiverse~\cite{hamkins:modal}. Switches are
statements~$\varphi$ such that~$\necessary(\possible\varphi \wedge
\possible\neg\varphi)$, while buttons are statements~$\varphi$ such
that~$\necessary\possible\necessary\varphi$; switches can be toggled on and off
like a light switch, while buttons once pressed cannot be unpressed:
\begin{enumerate}
\addtocounter{enumi}{10}
\item \emph{The law of excluded middle is a switch:} Everywhere it is the case
that somewhere~\textsc{lem} holds and somewhere it does not.
\item \emph{Being countable is a button:} For every set~$X$, everywhere it is
the case that somewhere (even proximally) it is the case that everywhere~$X$ is
countable.
\end{enumerate}

We argue that the proposed modal operators~$\necessary,\possible,\xpossible$ and more are natural extensions and refinements of the
familiar double-negation modality~$\neg\neg$ in constructive mathematics.

\paragraph{Related work.}
The idea of a mathematical multiverse is arguably basic to topos theory. We aim to present a more systematic study of the modal nature of the multiverse with a focus on applications in constructive mathematics. Specific precursors to XXX

In the airy reaches of classical set theory a related philosophy is put forward, with a number of striking results, concentrating on exploring the range of set-theoretic possibility~\cite{hamkins:multiverse,gitman-hamkins:multiverse,hamkins:constructibility,maddy-meadows:reconstruction,barton:forcing}.

\xxx{related work: Hamkins, Victoria Gitman, ... concentrates on exploring
range of set-theoretic possibility, obtaining beautiful results such as ...
Shawn Henry, Andreas Blass ...}

This text is set in the context of constructive mathematics.
Our definitions and results can be formalized in~\textsc{izf} or in the kind of language supported by toposes.

\paragraph{Acknowledgments.} We gratefully acknowledge numerous discussions and worthwhile input by Thierry Coquand, Matthias Hutzler, Ivan Di Liberti, Milly Maietti, Iosif Petrakis, Peter Schuster, and also the organizers and participants of the ItaCa Fest 2022 and the 2022 REDCOM workshop in Brixen, where this work was presented. XXX

@@

Henceworth, by the unqualified word ``topos'' we will mean ``Grothendieck topos
over the current base'', and the current base topos will be denoted~``$\Set$''.
The base topos might be the ``true category of sets'', assuming that this
concept is available in one's ontology, or also some other elementary topos
such as the free topos~\cite{lambek:1980intuitionist} or the ef{}fective topos~\cite{hyland:effective-topos}. The reader is reminded that the nature of the base topos can have profound consequences for the truths of the internal language.

The topos-theoretic multiverse of a base~$\B$ is the collection of Grothendieck
toposes over~$\B$ (which in set-theoretic foundations should more precisely be
formalized as the proper class of small sites in~$\B$).

\begin{remark}Perhaps arbitrary \emph{elementary} toposes over the base, corresponding to
possibly unbounded geometric morphisms~$\E \to \B$, or even arbitrary
fibrations/indexed categories over~$\B$ validating an appropriate form of the
axioms of elementary toposes, should also be
taken as part of the topos-theoretic multiverse. [XXX check fibrational characterization] The first foray into the
modal topos-theoretic multiverse outlined in this note
sticks to Grothendieck toposes for ease of
formalizability (``for every small site'' can be expressed also in the more
standard flavors of the internal topos language, while ``for every elementary
topos'' requires more elaborate versions); because the restricted multiverse
is already sufficiently rich for the intended applications; and because we have
a generalization to the predicative setting~\cite{crosilla:xxx} with arithmetic
universes in mind. Predicatively, not even the category of sets might be an
elementary topos, but the category of sets and categories of sheaves are still
arithmetic universes.
\end{remark}

@@

\begin{definition}A topos~$\E$ is \emph{positive} if and only if the unique geometric
morphism~$f : \E \to \Set$ is surjective (that is~$f^*$ reflects
isomorphisms).\end{definition}

@@

\begin{example}A necessary and sufficient criterion for the classifying topos of a geometric
theory~$\TT$ to be positive is that whenever~$\top \vdash \bigvee_{i \in I} \varphi_i$ modulo~$\TT$, then~$I$ is inhabited.\end{example}

\subsection{Overt toposes}

\begin{definition}A topos~$\E$ is \emph{overt} if and only if the unique geometric
morphism~$f : \E \to \Set$ is open (that is~$f^*$ preserves the interpretation of
bounded first-order formulas).\end{definition}

\begin{example}The topos of sheaves over a topological space is always overt. More
generally, the topos of sheaves over a locale~$X$ is overt iff there exists a
\emph{positivity predicate} on its frame of opens in the sense of~\cite{xxx}.
With~\textsc{lem}, every locale and indeed every topos is overt~\cite{xxx}.\end{example}

\begin{example}The spectrum of a ring~$A$ is overt iff every element of~$A$ is
nilpotent or not~\cite[Proposition~12.51]{blechschmidt:phd}.\end{example}

\begin{example}A sufficient criterion for the classifying topos of a geometric
theory~$\TT$ being overt is that the indexing sets of all disjunctions
appearing on the right hand side of turnstiles, in a normal form presentation
of~$\TT$, are inhabited~\cite[Proposition~V.3.2]{joyal-tierney:grothendieck}.\end{example}

@@

\xxx{properly define language}

@@

Similar inductive notions are used to reformulate other classical definitions
in a constructively more sensible way. For instance, the classical definition
of a binary relation~$R$ on a set~$X$ being \emph{almost-full} is ``every
infinite sequence~$\alpha : \NN \to X$ is \emph{good} in the sense that there
exist indices~$i < j$ such that~$\alpha(i) \mathrel{R} \alpha(j)$''.
For a constructive reformulation, we shift to finite approximations of
infinite sequences (finite lists of elements of~$X$) and define when such an
approximation is deemed good:\footnote{By~``$\sigma[n]$'', we mean the element
at position~$n$ of the finite list~$\sigma$. This notation is only meaningful
if the length of~$\sigma$ is at least~$n+1$. By~``$\sigma x$'' we mean the
enlarged list which has~$x$ as an extra element at its tail end, and by~``$[\,]$'' we
denote the empty list. In computer science practice, it is often more efficient
to prepend (``$x \sigma$'') instead of append, but this detail shall not
concern us here.}
\[ \mathsf{Good}(\sigma) \defeqv (\exists i<j\_ \sigma[i] \mathrel{R} \sigma[j]). \]
We then inductively generate a relation~``$P \,|\, \sigma$'' for monotonous
predicates~$P$ on finite lists expressing that no matter how the given finite
approximation~$\sigma$ evolves over time to a better approximation,
eventually~$P$ will hold, by the clauses
\[
  \infer{P \,|\, \sigma}{P(\sigma)}
  \qquad
  \infer{P \,|\, \sigma}{\forall x \? X\_\ P \,|\, \sigma x}
\]
and finally define that~$R$ is almost-full iff~``$\mathsf{Good} \,|\, []$''.
With this inductive definition, expected properties of the class of almost-full
relations such as stability under cartesian products (Dickman's lemma),
finite lists (Higman's lemma) or finitely-branching trees (Kruskal's theorem)
can all be constructively verified~\cite{xxx}.

A similar such definition has been proposed by Thierry Coquand, Henry Lombardi and Henrik Persson in commutative
algebra for expressing that a ring is Noetherian~\cite{coquand-persson:groebner,coquand-lombardi:krull,coquand:invariant}; the classical
definition ``every ascending chain of ideals stabilizes'' and also the more
meaningful and classically equivalent characterization as ``every ascending
chain of finitely generated ideals stalls''\footnote{A chain~$\aaa_0 \subseteq
\aaa_1 \subseteq \cdots$ \emph{stalls} iff for some index~$n \in \NN$,
$\aaa_{n+1} = \aaa_n$. We are grateful to Matthias Hutzler for proposing this
terminology.} are constructively too weak; firstly, without the axiom of
dependent choice we can often not construct such chains~\cite{richman:noetherian} (but only
``multi-valued chains'' as in~\cite[Section~3.9]{blechschmidt:phd}; but also
see~\cite[Section~4]{richman:noetherian}), and secondly, being able to inspect suitable inductive
witnesses enables us to prove the Hilbert basis theorem~\cite[Corollary~16]{coquand-persson:groebner}. Coquand, Lombardi and Persson
hence propose to call a ring \emph{Noetherian} if and only if~$\mathsf{Stalls}
\,|\, [\,]$, where~$\mathsf{Stalls}$ is the predicate on finite lists of finitely
generated ideals expressing xxx.

@@

\begin{proof}The classifying topos of the theory of an infinite sequence
over~$X$ can be presented as the topos of sheaves over the site given by the
partially ordered set of finite lists of elements of~$X$ with coverage given
by xxx (see Appendix xxx)~\cite[Example~4.3]{blechschmidt-schuster:reifying}, \cite[xxx]{joyal-tierny:xxx}. The Kripke--Joyal semantics states that the
statement ``$\alpha_0$ is good'', where~$\alpha_0$ is the generic infinite
sequence, holds in the classifying topos if and only if there is a
covering~$\U$ of~$[]$ such that for every open~$U \in \U$, xxx. This precisely
amounts to~$R$ being almost-full in the inductive sense.
\end{proof}

@@

\begin{proof}Similar as the proof of Proposition~\ref{prop:gen-good}. Details
for the variant of ``bad sets'' instead of ``infinite descending chains'' have
been developed (xxx:language) by Blass~\cite{blass:induction}.\end{proof}

@@

    \item (A ``Ramsey-style argument''.) The set~$I \defeq \{ n \in \NN \,|\, \neg\exists m > n\_ \alpha(n) \leq \alpha(m) \}$ is not in bijection with~$\NN$, as else the~$I$-extracted subsequence of~$\alpha$ would be an $X$-sequence which is not good. Hence, by \textsc{lem}, the set~$I$ is finite. Every index larger than all the indices in~$I$ is a suitable starting point for an infinite ascending chain~$\alpha(i_0) \leq \alpha(i_1) \leq \ldots$.\footnote{We can avoid dependent
    choice here by always picking the
    least possible next index.}

    \item Because~$Y$ is almost-full, the sequence~$\beta(i_0),\beta(i_1),\ldots$ is good, that is there exists a pair of indices~$n < m$ such that~$\beta(i_n) \leq \beta(i_m)$. As also~$\alpha(i_n) \leq \alpha(i_m)$, the sequence~$\gamma$ is good.
    % XXX: this only holds in the transitive case
  \end{enumerate}
\end{proof}

@@

\xxx{determine modal laws}

\xxx{more case studies}

\xxx{don't forget right adjoints}

\xxx{predicative}

\bibliographystyle{splncs04}
\bibliography{paper-modal-multiverse}

@@

simpler logical form
(every prime ideal vs. nilpotent)

cite https://www.cse.chalmers.se/~coquand/potential.pdf page 41

cite https://webspace.science.uu.nl/~ooste110/veldmanabstract.pdf
("perhaps" operator)

cite:
https://arxiv.org/pdf/2108.10259.pdf
https://arxiv.org/abs/2311.00581

cite:
Slides-What-if-your-potentialism-is-implicitly-actualist-Oxford-March-2024.pdf, slide 74 (Joel David Hamkins)

check:
https://www.youtube.com/watch?v=TpPQHS_wUsg, Giraud topos

cite:
https://arxiv.org/abs/2206.03169

because of representation theorem,
to say "for all Grothendieck toposes",
suffices to say "for all localic toposes".
Similarly for overt and positive

cite
https://arxiv.org/pdf/1712.04864
for internal language of toposes

show that inductive-wqo implies Berardi-wqo
show the theorem displayed at https://ncatlab.org/nlab/show/well-quasi-order

https://math.stackexchange.com/questions/4552108/does-the-inverse-image-part-of-a-geometric-morphism-preserve-transitive-set-obj
wf preserved

https://mathoverflow.net/questions/288789/constructive-proofs-of-existence-in-analysis-using-locales/290421?noredirect=1#comment1248291_290421
Simon suggests "proper and surjective" as an interesting class.

Completeness: If sequent holds for all models everywhere, then provable.
